---
title: "Practical Machine Learning - Course Project"
author: "Brian Hardin"
date: "4/9/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

The goal of the project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

## Project Datasets
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

## Load Packages
```{r, message = FALSE, warning = FALSE}
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(12345)
```

## Load Datasets
```{r}
# set URL
urltr <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urlts  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# download the data
train <- read.csv(url(urltr))
test  <- read.csv(url(urlts))

# partition datasets
intrain  <- createDataPartition(train$classe, p=0.7, list=FALSE)
traindata <- train[intrain, ]
testdata  <- train[-intrain, ]
```

```{r}
dim(traindata)
```
```{r}
dim(testdata)
```
Both created datasets have 160 variables. Each have variable sets that contain NA values, that can be removed with the cleaning procedures below. The Near Zero variance (NZV) variables are also removed and the ID variables as well.
```{r}
nzv <- nearZeroVar(traindata)
TrainDat <- traindata[, -nzv]
TestDat  <- testdata[, -nzv]
dim(TrainDat)
```

```{r}
dim(TestDat)
```

```{r}
# remove NA Variables
NARM   <- sapply(TrainDat, function(x) mean(is.na(x))) > 0.95
TrainDat <- TrainDat[, NARM==FALSE]
TestDat  <- TestDat[, NARM==FALSE]
dim(TrainDat)
```
```{r}
dim(TestDat)
```
```{r}
# remove id variables
TrainDat <- TrainDat[, -(1:5)]
TestDat  <- TestDat[, -(1:5)]
dim(TrainDat)
```

```{r}
dim(TestDat)
```
Now the dataset only includes 54 variables. 

## Correlation Analysis

```{r}
corMatrix <- cor(TrainDat[, -54])
corrplot(corMatrix, order = "FPC", method = "color", type = "lower", 
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```
The highly correlated variables are shown in dark colors in the graph above. To make an evem more compact analysis, a PCA (Principal Components Analysis) could be performed as pre-processing step to the datasets. 

## Prediction Model
Three methods will be applied to model the regressions (in the Train dataset) and the best one (with higher accuracy when applied to the Test dataset) will be used for the quiz predictions. The methods are: Random Forests, Decision Tree and Generalized Boosted Model, as described below.

```{r}
set.seed(12345)
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainDat, method="rf",
                          trControl=controlRF)
modFitRandForest$finalModel
```
```{r}
# prediction on Test dataset
predictRandForest <- predict(modFitRandForest, newdata=TestDat)
confMatRandForest <- confusionMatrix(predictRandForest, TestDat$classe)
confMatRandForest
```
```{r}
# plot matrix results
plot(confMatRandForest$table, col = confMatRandForest$byClass, 
     main = paste("Random Forest - Accuracy =",
                  round(confMatRandForest$overall['Accuracy'], 4)))
```
## Decision Trees

```{r}
# model fit
set.seed(12345)
modFitDecTree <- rpart(classe ~ ., data=TrainDat, method="class")
fancyRpartPlot(modFitDecTree)
```

```{r}
# prediction on Test dataset
predictDecTree <- predict(modFitDecTree, newdata=TestDat, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, TestDat$classe)
confMatDecTree
```
```{r}
# plot matrix results
plot(confMatDecTree$table, col = confMatDecTree$byClass, 
     main = paste("Decision Tree - Accuracy =",
                  round(confMatDecTree$overall['Accuracy'], 4)))
```

## Generalized Boosted Model

```{r, message = FALSE}
# model fit
set.seed(12345)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM  <- train(classe ~ ., data=TrainDat, method = "gbm",
                    trControl = controlGBM, verbose = FALSE)
modFitGBM$finalModel
```

```{r}
# prediction on Test dataset
predictGBM <- predict(modFitGBM, newdata=TestDat)
confMatGBM <- confusionMatrix(predictGBM, TestDat$classe)
confMatGBM
```

```{r}
# plot matrix results
plot(confMatGBM$table, col = confMatGBM$byClass, 
     main = paste("GBM - Accuracy =", round(confMatGBM$overall['Accuracy'], 4)))
```

## Applying the Selected Model to the Test Data

The accuracy of the 3 regression modeling methods above are:

Random Forest : 0.9963
Decision Tree : 0.7368
GBM : 0.9839
In that case, the Random Forest model will be applied to predict the 20 quiz results (testing dataset) as shown below.

```{r}
predictTEST <- predict(modFitRandForest, newdata=test)
predictTEST
```

